---
bibliography: bibtexlib.bib
output:
  html_document:
    css: tutorial.css
  word_document: default
---

```{r init, echo=FALSE}
library(knitr)
library(xtable)
library(tools)
library(knitcitations)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
source("functions.R")
gloss <- make_glossary_function()
ftf_def <- paste(
  "a text file to hold data in a tabular format, where cells are separated by a",
  "common character such as comma or tab. Common formats are csv and tsv."
)
```

# Data Formatting

GIGO is an acronym that stands for "Garbage In, Garbage Out", and this is often 
a mantra to remember when gathering data. The same goes for entering and storing
data in a spreadsheet. Often, the first point of error for any analysis is at 
the data entry stage. This section will show you the best practices of data 
storage and why this is important for your analyses, whether you are analyzing 
disease progress, fungicide resistance, or even bioinformatic data. Much of this
is based off of the Data Carpentry module [Using Spreadsheet Programs for
Scientific Data](http://www.datacarpentry.org/2015-03-09-ISI-CODATA/lessons/excel/ecology-examples/00-intro.html)
designed by Christie Bahlai and Aleksandra Pawlik. Please refer to the previous
link if you need any more information.

## Isn't Excel/OpenOffice/Google Sheets good enough?

**No** [@zeeberg2004mistaken]. 
![An example of gene names accidentally being converted to dates in Microsoft
Excel](images/excel_date_error.jpg){.big_window .knit_image .knit_border}

For data entry, Excel is a fantastic program because it's intuitive to enter data into cells, but Excel also reinforces a lot of bad habits that can make your data inaccessable to analyses down the line.

## How should I format my data?

It all depends on what your data is, but ultimately there are a few guidelines
or best practices to follow:

 - Observations in rows (note: 1 sample ≠ 1 observation)
 - Variables in columns (e.g. Treatment, Replicate, Response)
 - One datum per cell
 - Saved as a `r gloss("flat text file", ftf_def)` (e.g. `.csv`, `.tsv`, `.txt`)
 


## The good

Long or wide data in a text file 

## The bad

As above, but with color formatting to encode data. 

## The ugly

Separate tables within the spreadsheet, coded by color and stored only within
excel.

# Glossary

```{r glossary, results = 'asis', echo = FALSE}
gloss(display = TRUE)
```

# References