---
title: "The basics of R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NULL, prompt = TRUE)
library('printr')
source("functions.R")
gloss <- make_glossary_function()
vector_def <- "a sequence of data elements of the same type"
function_def <- "a self-contained set of code that can be used many times"
package_def <- "a set of functions and data sets that are bundled together for distribution and download"
library_def <- "a folder/directory on your computer where your R packages are stored"
prompt_def <- "a symbol that R uses to let you know it's ready for you to enter commands"
assign_def <- "assigns the result of whatever's to the right of the operator to whatever is on the left"
```

Introduction
------------

This is a quick introduction to R. It's not meant to be a comprehensive
tutorial, but rather as a small, sturdy foothold to get you started. We will
cover concepts such as `r gloss("vectors", vector_def)`, 
`r gloss("functions", function_def)`, and `r gloss("packages", package_def)`. 


Fancy Calculator
----------------

R is a programming language, but above all, it's a statistical programming 
language that's interactive. The basic usage of R is to use it like a
calculator.

When you open R, the first thing you'll see is the version, license and then the 
`r gloss('R prompt (>)', prompt_def)`:

```

R version 3.3.1 (2016-06-21) -- "Bug in Your Hair"
Copyright (C) 2016 The R Foundation for Statistical Computing
Platform: x86_64-apple-darwin13.4.0 (64-bit)

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

  Natural language support but running in an English locale

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

>
```

The R prompt tells us that it's ready for us to type commands. Unlike other
programming languages, it's designed to be interactive. As an example, we'll do
some simple calculations showing the relationship between 47 and 2. In your R
console, type `47 + 2` and then hit <kbd>Enter</kbd>. After that, type `47 * 2`
and hit <kbd>Enter</kbd>. You should see something like this:

```{r calculate_good_times}
47 + 2
47 * 2
```

Variables
---------

Unlike a calculator, you can save the results of your calculations into
variables by using the `r gloss("assignment operator (<-)", assign_def)`. This
looks like an arrow pointing the results of the calculation on the right to the
variable on the left.

```{r its_magic}
# 49 - 2 and 94/2 are both 47. It's magic!
its   <- 49 - 2
magic <- 94 / 2
```

> The # symbol is a "comment". R will ignore anything that comes after a
> comment, allowing you to write notes to yourself in your R script

Now that we saved the results into a variable, how do we look at it? All we have
to do is type the variable:

```{r magic_revealed}
its
magic
```

Variables can be used in calculations.

```{r multiple_magic}
oh  <- its + 2
oh
wow <- magic * 2
wow
```

You can even overwrite variables.

```{r written_magic}
oh <- "oh"
oh
wow <- "wow"
wow
```

But single variables are not all that you can do In R. Read more to find out
about vectors!

Vectors
-------

R can store more than single values. It has `r gloss("vectors", vector_def)`.
These are sequences of data that are all of the same type (i.e. integers,
decimals, text (characters), and logical values (TRUE/FALSE)). To construct a vector, you can use the `c()` function (more about functions later):

```{r vector_me_this}
some_numbers    <- c(pi, 47.5362, 3.50, 1.1111)
some_numbers    
some_integers   <- c(1, 1, 2, 3)
some_integers   
more_integers   <- 1:4 # same as c(1, 2, 3, 4), but a lot easier to type!
more_integers
some_characters <- c("work it", "harder", "make it", "better",  # You can wrap commands on
                     "do it", "faster", "makes us", "stronger") # multiple lines.
some_characters 
some_logic      <- c(TRUE, FALSE, TRUE, TRUE)
some_logic      
```


### Vectorization

You can think about these as being similar to a column in an excel spreadhseet.
For example, if you wanted to take column A and square each number, you would
start by setting cell **B1** to `=A1^2` and then you would drag that down to
apply to the entire column.

Math works with vectors the same way in R:

```{r v_add_o}
some_integers ^ 2
```

Again, if you tried to add two columns together in excel, you would type something like `=A1+B1` and then drag to apply to the column. R behaves the same way:

```{r v_drag}
some_integers + some_numbers
```

### Subsetting vectors


#### Using integers

Notice that each vector starts with `[1]` when you print it? This is telling you
what position this number is in the vector. You can get a specific element of a
vector by supplying a specific integer or sequence of integers:

```{r get_number}
some_numbers[2]
some_characters[c(1, 3, 5, 7)]
some_logic[1:3]
some_characters[some_integers] # you can also use other vectors!
```

Notice in the last example, we took our integer vector with the Fibbonacci
sequence and used it to subset the character vector.

#### Using logical values

We can also use logical values to subset a vector:

```{r truth_and_consequences}
some_numbers[some_logic]
```

This is a very powerful method of subsetting a vector because we can use logical comparison such as `>`, `<`, and `==`. We can get the same result as above by asking R to only return numbers less than ten:

```{r less_than_ten}
some_numbers # the second element is greater than 10.
some_numbers[some_numbers < 10]
```

Functions
---------

One other feature of R is the fact that it has 
`r gloss("functions", function_def)`. These are self-contained pieces of code
that can be run over and over. Functions can do anything from reading and
writing files, graphics, and calculations. A simple function is the `sum()` function:

```{r sum}
sum(some_integers)
```

Functions always take the form of `do_this(to_this_thing)`, so in the case
above, we are taking the sum of some integers.

Packages
--------

R would not be so widely used without `r gloss("packages", package_def)`. These 
are a set of functions, documentation, and data sets that can be distributed to 
anyone using R. If you can think of something you want R to do, there's probably
a package for that. Packages can be downloaded from online repositories such as
[CRAN](https://cran.r-project.org/) or [BioConductor](http://bioconductor.org/).

The easiest way to install a package is through your R console with the function
`install.packages()`:

```{r install_package, eval = FALSE}
install.packages("poppr", repos = "https://cran.rstudio.org")
```
```{r install_package_shadow, echo = FALSE}
message("
Installing package into '/Users/zhian/R'
(as 'lib' is unspecified)
trying URL 'https://cran.rstudio.org/bin/macosx/mavericks/contrib/3.3/poppr_2.2.0.tgz'
Content type 'application/x-gzip' length 2617496 bytes (2.5 MB)
==================================================
downloaded 2.5 MB
")
cat("
The downloaded binary packages are in
	/var/folders/qd/dpdhfsz12wb3c7wz0xdm6dbm0000gn/T//RtmpWCcCSI/downloaded_packages
")
```

This package is stored in your `r gloss("R library", library_def)`, which is a 
folder on your computer where your downloaded packages live. When you want to 
use the functions in a package, you load it with the `library()` function. This
function tells R to look in your library for a package and load it.

```{r}
library("poppr")
```


Getting help
-----------------------

One way that R shines above other languages is that R packages in CRAN are all
documented and easy to install. Help files are written in HTML and give the user
a brief overview of:

 - the purpose of a function
 - the parameters it takes
 - the output it yields
 - examples demonstrating its usage

To get help on any R function, type a question mark before the empty function. 
Here's an example of how to get help about the `gather()` function from the
*tidyr* package:

```{r}
library("tidyr") # The package with the gather() function.
?gather          # open the R documentation of the function gather()
```

If you want to run the examples, you can either copy and paste the commands to
your R console, or you can run them all with:

```{r, eval = FALSE}
example("gather", package = "tidyr")
```

Other ways of getting help:

```{r, eval = FALSE}
help(package = "tidyr")  # Get help for a package.
help("gather")           # Get help for the gather function
?gather                  # same as above
??multilocus             # Search for help that has the keyword 'multilocus' in all packages
```

Some packages include vignettes that can have different formats such as being
introductions, tutorials, or reference cards in PDF format. You can look at a
list of vignettes in all packages by typing:

```{r, eval=FALSE}
browseVignettes()                     # see vignettes from all packages
browseVignettes(package = 'poppr')    # see vignettes from a specific package.
```

and to look at a specific vignette you can type:

```{r, eval=FALSE}
vignette('mlg')
```

```{r, echo = FALSE}
detach("package:printr", unload = TRUE)
```

# Glossary

```{r glossary, results = 'asis', echo = FALSE}
gloss(display = TRUE)
```

